---
title: "Analysis of yeast mRNA and growth rate under nutrient limitation"
author: "Jay Hesselberth"
date: "February 25, 2016"
output: html_document
---

# Overview

In this document we will cover a previous analysis by David Robinson of some genomic data using tidy data analysis tools. I have paraphrased two of his posts into this combined document:

- http://varianceexplained.org/r/tidy-genomics/
- http://varianceexplained.org/r/tidy-genomics-broom/

## Load and tidy the data set

```{r init}
library(tidyverse)
library(cowplot)

url <- "http://varianceexplained.org/files/Brauer2008_DataSet1.tds"

nutrient_names <- c(G = "Glucose", L = "Leucine", P = "Phosphate",
                    S = "Sulfate", N = "Ammonia", U = "Uracil")

cleaned_data <- read_tsv(url) %>%
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), sep = "\\|\\|") %>%
  mutate_each(funs(trimws), name:systematic_name) %>%
  select(-number, -GID, -YORF, -GWEIGHT) %>%
  gather(sample, expression, G0.05:U0.3) %>%
  separate(sample, c("nutrient", "rate"), sep = 1, convert = TRUE) %>%
  mutate(nutrient = plyr::revalue(nutrient, nutrient_names)) %>%
  filter(!is.na(expression), systematic_name != "")
```

## Explore the data using plots

Look at a few genes involved in leucine biosynthesis.

```{r plot_1}
cleaned_data %>%
  filter(BP == "leucine biosynthesis") %>%
  ggplot(aes(rate, expression, color = nutrient)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~name + systematic_name) + 
  theme_cowplot()
```

This block illustrates how to define and use a function in R to reuse the plotting code.

```{r plot_2}
plot_expression_data <- function(expression_data) {
  ggplot(expression_data, aes(rate, expression, color = nutrient)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    scale_color_brewer(palette = "Set1") +
    facet_wrap(~name + systematic_name, scales = "free_y") +
    theme_cowplot()
}

cleaned_data %>%
  filter(BP == "leucine biosynthesis") %>%
  plot_expression_data()

cleaned_data %>%
  filter(BP == "cell wall organization and biogenesis") %>%
  group_by(name) %>%
  slice(1:5) %>%
  plot_expression_data()
```

### Exercises

- Make a similar plot for uracil biosynthesis genes.

```{r}
cleaned_data %>%
  filter(grepl('URA', name)) %>% plot_expression_data()
```

- What happens if you leave out `scales = "free_y"` from the `facet_wrap` call?

```{r}
```

## Linear modeling

```{r plot_3}
LEU1_leucine <- cleaned_data %>%
  filter(name == "LEU1", nutrient == "Leucine")

ggplot(LEU1_leucine, aes(rate, expression)) +
  geom_point() + geom_smooth(method = 'lm', se = FALSE)
```

```{r plot_4}
mod <- lm(expression ~ rate, LEU1_leucine)
summary(mod)
```

```{r plot_5}
library(broom)
tidy(mod)
```

```{r many_models}
linear_models <- cleaned_data %>%
  group_by(name, systematic_name, nutrient) %>%
  do(tidy(lm(expression ~ rate, .)))

linear_models
```

Now we'll filter out the intercept term from the models. Note the use of the `ungroup()` function below.

```{r intercepts}
intercept_terms <- linear_models %>%
  ungroup() %>%
  filter(term == "(Intercept)")

intercept_terms
```

```{r center_intercepts}
centered_intercepts <- intercept_terms %>%
  group_by(systematic_name) %>%
  mutate(centered_intercept = estimate - mean(estimate)) %>%
  ungroup()

top_intercept <- centered_intercepts %>%
  top_n(20, centered_intercept)

cleaned_data %>%
  semi_join(top_intercept, by = "systematic_name") %>%
  plot_expression_data()
```

### Slope terms

```{r slope_terms}
slope_terms <- linear_models %>%
  ungroup() %>%
  filter(term == "rate", !is.na(p.value))

ggplot(slope_terms, aes(p.value)) +
  geom_histogram(binwidth = .05) +
  facet_wrap(~nutrient)

slope_terms <- slope_terms %>%
  mutate(q.value = p.adjust(p.value))

slope_terms %>%
  group_by(nutrient) %>%
  summarize(significant = sum(q.value < .01)) %>%
  arrange(-significant)
```

### Exercises

- What happens if you leave out the `ungroup()` when you filter for intercepts?

```{r}
```

- What does `semi_join()` do? What happens if you use the other `join` verbs (e.g., `left_join`, `anti_join`)?

```{r}
```
